String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-08-18 2023-1-19/sections/preliminaries.tex
setta2023

We write $\Nat$ and $\Int$ for the sets of natural and integer numbers. For $n \in \Nat$ with $n \ge 1$, $[n]$ denotes $\{1, \ldots, n\}$; for $m,n \in \Nat$ with $m \le n$,  $[m, n]$ denotes $\{ i \in \Nat \mid m \le i \le n \}$. Throughout the paper, $\Sigma$ is a finite alphabet, ranged over by $a,b,\ldots$.

For a function $f$ from $X$ to $Y$ and $X' \subseteq X$, we use $\prj_{X'}(f)$ to denote the restriction (aka projection) of $f$ to $X'$, that is, $\prj_{X'}(f)$ is a function from $X'$ to $Y$, and $\prj_{X'}(f)(x') = f(x')$ for each $x' \in X'$.

@@

It is well-known that finite automata capture regular languages. 
Moreover, the class of regular languages is closed under union, intersection, concatenation, Kleene star, complement, and language difference \cite{HU79}. In the sequel, we recall the constructions of NFA corresponding to these language operations. Let $\NFA_1 = (Q_1, \Sigma, \delta_1, I_1, F_1)$ and $\NFA_2 = (Q_2, \Sigma, \delta_2, I_2, F_2)$ be two NFA such that $Q_1 \cap Q_2 = \emptyset$. 
\begin{itemize}
\item The union of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \cup \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', I', F')$, where $Q'= Q_1 \cup Q_2$, $\delta' = \delta_1 \cup \delta_2$, $I' = I_1 \cup I_2$, and $F'=F_1 \cup F_2$. 
%
\item The intersection of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \cap \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', I', F')$, where $Q' = Q_1 \times Q_2$, $I' = I_1 \times I_2$, $F' = F_1 \times F_2$, and $\delta' = \{((q_1, q_2), a, (q'_1, q'_2)) \mid (q_1, a, q'_1) \in \delta_1, (q_2, a, q'_2) \in \delta_2\}$.
%
\item The concatenation of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \concat \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', I', F')$, where $Q' = Q_1 \cup Q_2$, $I' = I_1$, $\delta' = \delta_1 \cup \delta_2 \cup \{(q_1, a, q_2) \mid q_1 \in F_1, \exists q' \in I_2.\ (q', a, q_2) \in \delta_2\}$, moreover, if $I_2 \cap F_2 \neq \emptyset$, then $F'= F_1 \cup F_2$, otherwise, $F'= F_2$.
%
\item The Kleene star of $\NFA_1$, denoted by $\NFA_1^*$, is defined as $(Q'_1, \Sigma, \delta'_1, I'_1, F'_1)$, where $Q'_1 = Q_1 \cup \{q_0\}$ with $q_0 \not \in Q_1$, $I'_1 = \{q_0\}$, $F'_1= F_1 \cup \{q_0\}$, $\delta'_1 = \delta_1 \cup \{(q_0, a, q'_1)  \mid \exists q'_0 \in I_1.\ (q'_0, a, q'_1) \in \delta_1\} \cup \{(q_1, a, q'_1) \mid q_1 \in F_1, \exists q'_0 \in I_1.\ (q'_0, a, q'_1) \in \delta_1\}$.
%
\item The complement of $\NFA_1$, denoted by $\overline{\NFA_1}$, is defined as $(Q'_1, \Sigma, \delta'_1, q'_0, Q'_1 \setminus F'_1)$, where  $(Q'_1, \Sigma, \delta'_1, q'_0, F'_1)$ is the DFA obtained from $\NFA_1$ by subset construction.  
%
\item The language difference of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \setminus \NFA_2$, is defined as $\NFA_1 \cap \overline{\NFA_2}$.
\end{itemize}

@@

\medskip
\noindent 
\emph{Linear integer arithmetic and Parikh images.}
We will also use standard quantifier-free/existential \emph{linear integer arithmetic} (LIA) formulas, which typically range over by $\phi, \varphi$. 
For a set $\mathfrak{X}$ of variables, we use $\Phi(\mathfrak{X})$ to denote the set of quantifier-free LIA formulas whose free variables are from $\mathfrak{X}$. 
For convenience, we use $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{\anivar_1, \cdots, \anivar_k\}$. For an LIA formula $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ and $\myvec{t} = (t_1, \cdots, t_k)$ be a tuple of integer terms, we use $\varphi[\myvec{t}/\myvec{\anivar}]$ to denote the formula obtained by replacing (simultaneously) $\anivar_i$ with $t_i$ for every $i \in [k]$.
%For convenience, we assume a set $X$ of variables is a vector. 
%, which is essentially Presburger

At last, we recall a result about Parikh images of NFA. 
For each $a \in \Sigma$, let $\anivarz_a$ be an integer variable. Let $\fZ_\Sigma$ denote the set of integer variables $\anivarz_a$ for $a \in \Sigma$. 
Let $\NFA$ be an NFA over the alphabet $\Sigma$. Then 
we say that an LIA formula $\varphi(\fZ_\Sigma)$ defines the Parikh image of $\NFA$, if $\{\eta: \Sigma \rightarrow \Nat \mid \varphi[(\eta(a) /\anivarz_a)_{a \in \Sigma}] \mbox{ is true}\} = \parikh(\NFA)$.

\begin{theorem}[\cite{SSMH04}]\label{thm-nfa-parikh}
Given an NFA $\NFA$,  an existential LIA formula $\varphi_\NFA(\fZ_\Sigma)$ can be computed in linear time that defines the Parikh image of $\NFA$.
\end{theorem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\hide{
A finite \emph{alphabet} $\Sigma$ is the set of all \emph{letters}. A
\emph{string} (or \emph{word}) is a finite sequence of letters from $\Sigma$. $\Sigma^*$ is the set of strings over $\Sigma$. $\epsilon$ is the empty string. $L$ is grammar. A language $\lan(L)$ is a set of words generated by $L$. $\mathbb{N}$ is the set of natural numbers and $\mathbb{Z}$ is the set of integer numbers. We use $a,
  b,\cdots$ to denote the constant letters in $\Sigma$, $u, v,\cdots$ to denote constant
string, $x, y,\cdots$ to denote variable string, $m,n,\cdots$ to
denote integer constant, and $i,j\cdots$ to denote integer variable. For vector, we use $\myvec{v}$ to denote the vector of integer constant, $m_n$ to denote the vector $(m,\cdots, m)$ with length $n$, $\myvec{v}[i]$ to denote the integer value of $\myvec{v}$ at position $i$, $\myvec{v_1}\cdot\myvec{v_2}$ to denote the concatenation of $\myvec{v_1}$ and $\myvec{v_2}$, $R$ to denote the vector of registers, $R_1 \cap R_2$ to denote the same registers in $R_1$ and $R_2$, and $|\myvec{v}|$ or $|R|$ to denote the length of the vector.
}

>>>>

We write $\Nat$ and $\Int$ for the sets of natural and integer numbers. For $n \in \Nat$ with $n \ge 1$, $[n]$ denotes $\{1, \ldots, n\}$; for $m,n \in \Nat$ with $m \le n$,  $[m, n]$ denotes $\{ i \in \Nat \mid m \le i \le n \}$. Throughout the paper, $\Sigma$ is a finite alphabet, ranged over $a,b,\ldots$.

For a function $f$ from $X$ to $Y$ and $X' \subseteq X$, we use $\prj_{X'}(f)$ to denote the restriction (aka projection) of $f$ to $X'$, that is, $\prj_{X'}(f)$ is the function from $X'$ to $Y$, and $\prj_{X'}(f)(x') = f(x')$ for each $x' \in X'$.

@@

It is well-known that finite automata capture regular languages. 
Moreover, the class of regular languages is closed under union, intersection, concatenation, Kleene star, complement, and language difference \cite{HU79}. The constructions of NFA corresponding to these language operations are list in \ref{appendix:cefa}
% In the sequel, we recall the constructions of NFA corresponding to these language operations. 
% Let $\NFA_1 = (Q_1, \Sigma, \delta_1, I_1, F_1)$ and $\NFA_2 = (Q_2, \Sigma, \delta_2, I_2, F_2)$ be two NFA such that $Q_1 \cap Q_2 = \emptyset$. 
% \begin{itemize}
% \item The union of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \cup \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', I', F')$, where $Q'= Q_1 \cup Q_2$, $\delta' = \delta_1 \cup \delta_2$, $I' = I_1 \cup I_2$, and $F'=F_1 \cup F_2$. 
% %
% \item The intersection of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \cap \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', I', F')$, where $Q' = Q_1 \times Q_2$, $I' = I_1 \times I_2$, $F' = F_1 \times F_2$, and $\delta' = \{((q_1, q_2), a, (q'_1, q'_2)) \mid (q_1, a, q'_1) \in \delta_1, (q_2, a, q'_2) \in \delta_2\}$.
% %
% \item The concatenation of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \concat \NFA_2$, is defined as the tuple $(Q', \Sigma, \delta', I', F')$, where $Q' = Q_1 \cup Q_2$, $I' = I_1$, $\delta' = \delta_1 \cup \delta_2 \cup \{(q_1, a, q_2) \mid q_1 \in F_1, \exists q' \in I_2.\ (q', a, q_2) \in \delta_2\}$, moreover, if $I_2 \cap F_2 \neq \emptyset$, then $F'= F_1 \cup F_2$, otherwise, $F'= F_2$.
% %
% \item The Kleene star of $\NFA_1$, denoted by $\NFA_1^*$, is defined as $(Q'_1, \Sigma, \delta'_1, I'_1, F'_1)$, where $Q'_1 = Q_1 \cup \{q_0\}$ with $q_0 \not \in Q_1$, $I'_1 = \{q_0\}$, $F'_1= F_1 \cup \{q_0\}$, $\delta'_1 = \delta_1 \cup \{(q_0, a, q'_1)  \mid \exists q'_0 \in I_1.\ (q'_0, a, q'_1) \in \delta_1\} \cup \{(q_1, a, q'_1) \mid q_1 \in F_1, \exists q'_0 \in I_1.\ (q'_0, a, q'_1) \in \delta_1\}$.
% %
% \item The complement of $\NFA_1$, denoted by $\overline{\NFA_1}$, is defined as $(Q'_1, \Sigma, \delta'_1, q'_0, Q'_1 \setminus F'_1)$, where  $(Q'_1, \Sigma, \delta'_1, q'_0, F'_1)$ is the DFA obtained from $\NFA_1$ by subset construction.  
% %
% \item The language difference of $\NFA_1$ and $\NFA_2$, denoted by $\NFA_1 \setminus \NFA_2$, is defined as $\NFA_1 \cap \overline{\NFA_2}$.
% \end{itemize}

@@

\medskip
\noindent 
\emph{Linear integer arithmetic and Parikh images.}
We use standard existential \emph{linear integer arithmetic} (LIA) formulas, which
 typically range over $\psi, \varphi, \Phi, \alpha$. 
For a set $\mathfrak{X}$ of variables, we use $\psi/\varphi/\Phi/\alpha(\mathfrak{X})$ to denote the set of existential LIA formulas whose free variables are from $\mathfrak{X}$. 
For example, we use $\varphi(\vec{\anivar})$ with $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ to denote an LIA formula $\varphi$ whose free variables are from $\{\anivar_1, \cdots, \anivar_k\}$. For an LIA formula $\varphi(\vec{\anivar})$, we use $\varphi[\myvec{t}/\myvec{\anivar}]$ to denote the formula obtained by replacing (simultaneously) $\anivar_i$ with $t_i$ for every $i \in [k]$ where $\myvec{\anivar} = (\anivar_1, \cdots, \anivar_k)$ and $\myvec{t} = (t_1, \cdots, t_k)$ are tuples of integer terms.
%For convenience, we assume a set $X$ of variables is a vector. 
%, which is essentially Presburger

At last, we recall the result about Parikh images of NFA. 
For each $a \in \Sigma$, let $\anivarz_a$ be an integer variable. Let $\fZ_\Sigma$ denote the set of integer variables $\anivarz_a$ for $a \in \Sigma$. 
Let $\NFA$ be an NFA over the alphabet $\Sigma$. Then 
we say that an LIA formula $\psi(\fZ_\Sigma)$ defines the Parikh image of $\NFA$, if $\{\eta: \Sigma \rightarrow \Nat \mid \psi[(\eta(a) /\anivarz_a)_{a \in \Sigma}] \mbox{ is true}\} = \parikh(\NFA)$.

\begin{theorem}[\cite{SSMH04}]\label{thm-nfa-parikh}
Given an NFA $\NFA$,  an existential LIA formula $\psi_\NFA(\fZ_\Sigma)$ can be computed in linear time that defines the Parikh image of $\NFA$.
\end{theorem}
\vspace{-3mm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\hide{
A finite \emph{alphabet} $\Sigma$ is the set of all \emph{letters}. A
\emph{string} (or \emph{word}) is a finite sequence of letters from $\Sigma$. $\Sigma^*$ is the set of strings over $\Sigma$. $\epsilon$ is the empty string. $L$ is grammar. A language $\lan(L)$ is a set of words generated by $L$. $\mathbb{N}$ is the set of natural numbers and $\mathbb{Z}$ is the set of integer numbers. We use $a,
  b,\cdots$ to denote the constant letters in $\Sigma$, $u, v,\cdots$ to denote constant
string, $x, y,\cdots$ to denote variable string, $m,n,\cdots$ to
denote integer constant, and $i,j\cdots$ to denote integer variable. For vector, we use $\myvec{v}$ to denote the vector of integer constant, $m_n$ to denote the vector $(m,\cdots, m)$ with length $n$, $\myvec{v}[i]$ to denote the integer value of $\myvec{v}$ at position $i$, $\myvec{v_1}\cdot\myvec{v_2}$ to denote the concatenation of $\myvec{v_1}$ and $\myvec{v_2}$, $R$ to denote the vector of registers, $R_1 \cap R_2$ to denote the same registers in $R_1$ and $R_2$, and $|\myvec{v}|$ or $|R|$ to denote the length of the vector.
}

### 2023-08-18 2023-1-19/sections/recl.tex
setta2023

%\subsubsection{Syntax}
In the sequel, we define string constraints with regex-counting and string-length functions (abbreviated as RECL). The syntax of RECL is defined by the rules in Figure~\ref{fig:syntax}, where $x$ is a string variable, $\anivar$ is an integer variable, $a$ is a character from an alphabet $\Sigma$, and $m, n$ are integer constants. 
A RECL formula $\varphi$ is a conjunction of atomic formulas of the form $x \in \regex$ or $t_1\ \op\ t_2$, where $\regex$ is a regular expression,  
$t_1$ and $t_2$ are integer terms, and $\op \in \{=, \neq, \le, \ge, <, >\}$. Atomic formulas of the form $x \in \regex$ are called \emph{regular membership} constraints, and atomic formulas of the form $t_1\ \op\ t_2$ are called \emph{length} constraints. 
%
A regular expression $\regex$ is built from $\emptyset$, the empty string $\epsilon$, and characters $a$ by using concatenation  $\cdot$, union $+$, Kleene star $^*$, intersection $\cap$, complement $\bar{\mbox{ }}$, difference $\setminus$, counting $^{\{m,n\}}$ or $^{\{m,\infty\}}$. An integer term is built from constants $n$, variables $\anivar$, and string lengths $|x|$, by operators $+$ and $-$.
% 
%is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{figure}[h]
\vspace{-3mm}
  \centering
%  \begin{tabular}{l r}
$ \begin{array}{l l l r}
    \varphi & ::= & x\in \regex \mid t_1\  \op\ t_2 \mid  \varphi \wedge \varphi                                              & \mbox{formulas}            \\
    \regex &::= & \emptyset \mid \epsilon \mid a \mid \regex\cdot \regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \overline{\regex} \mid \regex \setminus \regex \mid \regex^{\{m,n\}} \mid \regex^{\{m,\infty\}} \mid (e) & \mbox{regexes} \\
    t &::= & n \mid \anivar \mid  |x| \mid t - t \mid t + t                                                                    & \mbox{integer terms}
    \end{array}
  $
%  \end{tabular}
  \caption{Syntax of RECL }\label{fig:syntax}
\vspace{-3mm}
\end{figure}
Moreover, for $S \subseteq \Sigma$ with $S = \{a_1, \cdots, a_k\}$, we use $S$ as an abbreviation of $a_1 + \cdots + a_k$.

>>>>

%\subsubsection{Syntax}
In the sequel, we define the string constraints with regex-counting and string-length functions, i.e., \textbf{RE}gex-\textbf{C}ounting \textbf{L}ogic (abbreviated as RECL). The syntax of RECL is defined by the rules in Figure~\ref{fig:syntax}, where $x$ is a string variable, $\anivar$ is an integer variable, $a$ is a character from an alphabet $\Sigma$, and $m, n$ are integer constants. 
A RECL formula $\varphi$ is a conjunction of atomic formulas of the form $x \in \regex$ or $t_1\ \op\ t_2$, where $\regex$ is a regular expression,  
$t_1$ and $t_2$ are integer terms, and $\op \in \{=, \neq, \le, \ge, <, >\}$. Atomic formulas of the form $x \in \regex$ are called \emph{regular membership} constraints, and atomic formulas of the form $t_1\ \op\ t_2$ are called \emph{length} constraints. 
%
A regular expression $\regex$ is built from $\emptyset$, the empty string $\epsilon$, and the character $a$ by using concatenation  $\cdot$, union $+$, Kleene star $^*$, intersection $\cap$, complement $\bar{\mbox{ }}$, difference $\setminus$, counting $^{\{m,n\}}$ or $^{\{m,\infty\}}$. An integer term is built from constants $n$, variables $\anivar$, and string lengths $|x|$ by operators $+$ and $-$.
% 
%is a quantifier-free formula that can be a regular membership $x\in\regex$, a quantifier-free Presburger formula $\alpha_1 \leq \alpha_2$, a negation of a formula, or a disjunction of two formulas. $\alpha$ is an integer term which can be an integer constant $m$, integer variable $i$, the length of the string term $|s|$, the minus of an integer term, and the plus of two integer terms. The regular expression $\regex$ is built on empty string $\epsilon$, constant letter $a\in \Sigma$. The supported regex operations involve concatenation $\cdot$, disjunction $+$, intersection $\times$, closure $*$, complement $C$, and bounded repetition $\{m, n\}$. We use $term(\varphi)$ to denote the set of terms and $strvar(\varphi)$ to denote the string variables occurring in $\varphi$. The \emph{literals} of $\varphi$ are $x\in \regex$ and $\alpha_1 \leq \alpha_2$. $x\in \regex$ is called \emph{regular literals} and $\alpha_1 \leq \alpha_2$ is called \emph{linear literals}.
\begin{figure}[h]
  \centering
%  \begin{tabular}{l r}
$ \begin{array}{l l l r}
    \varphi & ::= & x\in \regex \mid t_1\  \op\ t_2 \mid  \varphi \wedge \varphi                                              & \mbox{formulas}            \\
    \regex &::= & \emptyset \mid \epsilon \mid a \mid \regex\cdot \regex \mid \regex+\regex \mid \regex^* \mid \regex \cap \regex \mid \overline{\regex} \mid \regex \setminus \regex \mid \regex^{\{m,n\}} \mid \regex^{\{m,\infty\}} \mid (e) & \mbox{regexes} \\
    t &::= & n \mid \anivar \mid  |x| \mid t - t \mid t + t                                                                    & \mbox{integer terms}
    \end{array}
  $
%  \end{tabular}
  \caption{Syntax of RECL }\label{fig:syntax}
\vspace{-4mm}
\end{figure}
Moreover, for $S \subseteq \Sigma$ with $S = \{a_1, \cdots, a_k\}$, we use $S$ as an abbreviation of $a_1 + \cdots + a_k$.
